### 2022-12-14 chapter1/chapter1.tex
Updated trimming method text with new training numbers and clearer descriptions

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM. States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first. Because the posterior probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.95 was used to define a ``seed'' region. The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected. The outward expansion halted when the probability or its derivative was below 0.01 and 0.001, respectively. The inward expansion halted when the derivative was below 0.001. The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns. The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.

% TODO: Update training numbers, cutoffs, and add additional state 3 check

Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method. First, the mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step. This ensured that any regions which were intermediate between state 2 and 3 were included. Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.95. The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method. However, the initial seeds were defined with a cutoff of 0.95, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.5. These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.

>>>>

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the deletion phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM. Regions with a high probability of state 2 or 3 are candidates for trimming. However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changes most abruptly. Instead, the following algorithm was used. First, a high cutoff defines a ``seed'' region. The left and right endpoints of the seed are then expanded both inwards and outwards to define two intervals from which boundaries are selected. The outward expansion halts when the probability or its derivative is below two distinct thresholds, respectively. The inward expansion halts when the derivative is below a different threshold. The left and right boundaries are chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns. The gap profile is calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences. By combining where the model’s confidence changes rapidly with the observed change in the gap pattern, this method generally selects reasonable boundaries.

 States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first. Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method. First, the probability of state 3 of columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions. Regions were then defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively. The outer and inner derivative cutoffs were both 0.001. The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step. This ensured that any regions which were intermediate between state 2 and 3 were included. Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.75. The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method. However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05. These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.
